### 2018-02-19 submission/manuscript.tex
update background

Colorectal cancer (CRC) is a growing world-wide health problem in which
the microbiota has been purported to play an active role in disease
pathogenesis {[}1,2{]}. Numerous studies have shown the importance of
both individual microbes {[}3--7{]} and the overall community
{[}8--10{]} in tumorgenesis using mouse models of CRC. There have also
been numerous case-control studies investigating the microbiota in the
formation of both adenoma and carcinoma. A recent meta-analysis
investigated whether specific biomarkers could be consistently
identified using multiple data sets {[}11{]}. This meta-analysis focused
on identifying microbial signatures of tumors (biomarkers) but did so on
a small total number of individuals and only investigated stool. This
present meta-analysis addresses some of these major shortcomings.

Although there has been an intense focus on microbiota-based biomarker
discovery for tumors, the number of candidate genera seem to be endless.
Some studies point towards mouth-associated genera such as
\emph{Fusobacterium}, \emph{Peptostreptococcus}, \emph{Parvimonas}, and
\emph{Porphyromonas} as key enriched genera {[}6,12--18{]}. Yet, even in
these studies, mouth-associated genera are far from the only microbes
identified to be associated with tumors. These other genera include, but
are not limited to, \emph{Providencia}, \emph{Mogibacterium},
\emph{Enterococcus}, \emph{Escherichia/Shigella}, \emph{Klebsiella}, and
\emph{Streptococcus} {[}14--16{]}. In fact, there is good \emph{in vivo}
evidence that \emph{Escherichia/Shigella} and \emph{Streptococcus} can
be important in the pathogenesis of CRC {[}5,19,20{]}. Other studies
have also identified \emph{Akkermansia muciniphila} and
\emph{Bacteroides fragilis} as potential markers of CRC with good
mechanistic studies for the latter {[}15,21,22{]}. A recent
meta-analysis confirmed the correlations of certain mouth-associated
genera and \emph{Akkermansia muciniphila} with carcinoma {[}11{]}.
However, the sample size (n = 509) is equal to or less than some of the
more recent individual studies investigating the microbiota and
colorectal tumors, making it hard to know how extrapolatable these
findings are. That particular meta-analysis also added more potential
microbial associations to both carcinoma (\emph{Pantoea agglomerans}
\emph{Ruminococcus}, \emph{Lactobacillus}) and adenoma
(\emph{Prevotella}, \emph{Methanosphaera}, \emph{Succinovibrio},
\emph{Haemophilus parainfluenzae}, \emph{Ruminococcus},
\emph{Lactobacillus}) stages of disease that need to be investigated
further, since a number of these genera have been found to be enriched
in controls and not disease {[}13,16,17{]}. Additionally, genera like
\emph{Roseburia} have been found in some studies to be increased in
tumors but in others to either be decreased or have no difference
{[}15,18,23,24{]}.

Most of these studies have focused on individuals with carcinomas but
associations with the adenoma stage of disease are not any clearer at
identifying candidate genera correlated with these earlier tumors.
Groups focusing on broad scale community metrics have found that
measures such as richness are decreased in those with adenomas versus
controls. Other studies have identified \emph{Lactococcus},
\emph{Pseudomonas}, \emph{Acidovorax}, \emph{Cloacibacterium},
\emph{Helicobacter}, \emph{Lactobacillus}, \emph{Bilophila},
\emph{Desulfovibrio}, and \emph{Mogibacterium} to be increased in those
with adenoma tumors {[}25--27{]}. Additionally, based on these studies
mentioned, there seems to be very few common genera that are associated
with both adenoma and carcinoma tumors, with \emph{Lactobacillus} being
one of the few commonalities.

Targeting the identification of tumor microbial biomarkers within stool
seems logical since it offers an easy and cost-effective way to stratify
risk of disease. The current gold standard for diagnosis, a colonoscopy,
can be time-consuming and is not without risk of complications. Although
stool represents an easy and less invasive way to assess risk, it is not
clear how well this sample reflects adenoma- and carcinoma- associated
microbial communities. Some studies have tried to assess this in health
and disease but are limited by their sample size {[}18,28{]}. Sampling
the microbiota directly associated with colon tissue may provide clearer
answers but is not without their own limitations. After the colonoscopy
bowel prep the bacterial community sampled may reflect the better
adhered microbiota versus the resident community. Additionally, these
samples contain more host DNA, potentially limiting the types of
analysis that can be done. It is well known that low biomass samples can
be very difficult to work with and results can be study dependent due to
the randomness of contamination {[}29{]}.

In comparison to the previous meta-analysis, this study significantly
increases the total stool samples investigated, re-examines important
genera across adenoma and carcinoma across study, and examines
differences and similarities between stool and tissue microbiota in the
context of colorectal tumors. Importantly, this analysis and approach
could provide valuable insights into the common genera that are both
protective and detrimental in individuals with adenoma or carcinoma and
whether broad bacterial community measurements can account for these
changes that were not provided by earlier meta-analysis studies
{[}11{]}.

Using both feces (n = 1737) and colon tissues (492 samples from 350
individuals) totaling over 2229 total samples across 14 studies
{[}12--18,21,23--27,30{]} {[}Table 1 \& 2{]}, we expand both the breadth
and scope of the previous meta-analysis to investigate whether the
bacterial community or specific members are more important risk factors
for both adenoma and carcinoma stages of disease. To accomplish this we
first assessed whether bacterial diversity changes throughout disease
(control to adenoma to carcinoma) and if it results in an increased odds
ratio (OR) for individuals to have either an adenoma or carcinoma. We
then assessed what genera, if any, increase or decrease the OR of an
individual to have an adenoma or carcinoma. Next, using Random Forest
models, we analyzed whether the full community or only the combined top
5 increased and top 5 decreased OR genera resulted in better model
classification, based on the area under the curve (AUC). Finally, we
also examined at what effect and sample size the studies used were
powered for and the sample size needed to get to the traditionally
accepted 80\% power. Our results from these analyses suggests that the
bacterial community changes as disease severity worsens, that specific
members are important for disease classification, and that many of the
individual studies are underpowered for assessing small effect sizes.

@@

@@

One important caveat to this study is that even though genera associated
with certain species such as \emph{Bacteroides fragilis} and
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} were not
identified, it does not necessarily mean that these specific species are
not important in human CRC {[}20,22{]}. Since we are limited in our
aggregation of the data to the genus level, it is not possible to
clearly delineate which species are contributing to overall disease
progression. Our observations are not inconsistent with the previous
literature on either \emph{Bacteroides fragilis} or \emph{Streptococcus
gallolyticus} subsp. \emph{gallolyticus}. As an example, the stool-based
full community models consistently identified the genus
\emph{Bacteroides}, as well as OTUs that classified as
\emph{Bacteroides}, to be important model components across studies.
This suggests that even though \emph{Bacteroides} may not increase the
OR of individuals having an adenoma or carcinoma and may not vary in
relative abundance, like \emph{Fusobacterium}, it is still important in
CRC. Additionally, \emph{Streptococcus gallolyticus} subsp.
\emph{gallolyticus} is a mouth-associated microbe, and the results from
this study suggest that regardless of sample type, mouth-associated
genera are commonly associated with an increased OR for individuals to
have a carcinoma tumor.

@@

\textbf{\emph{Obtaining Data Sets:}} The studies used for this
meta-analysis were identified through the review articles written by
Keku, \emph{et al.} and Vogtmann, \emph{et al.} {[}31,32{]} and
additional studies not mentioned in the reviews were obtained based on
the authors' knowledge of the literature. Studies that used tissue or
feces as their sample source for 454 or Illumina 16S rRNA gene
sequencing analysis and had data sets with sequences available for
analysis were included. Some studies were excluded because they did not
have publicly available sequences or did not have metadata in which the
authors were able to share. After these filtering steps, the following
studies remained: Ahn, \emph{et al.} {[}12{]}, Baxter, \emph{et al.}
{[}13{]}, Brim, \emph{et al.} {[}30{]}, Burns, \emph{et al.} {[}16{]},
Chen, \emph{et al.} {[}14{]}, Dejea, \emph{et al.} {[}24{]}, Flemer,
\emph{et al.} {[}18{]}, Geng, \emph{et al.} {[}23{]}, Hale, \emph{et
al.} {[}27{]}, Kostic, \emph{et al.} {[}33{]}, Lu, \emph{et al.}
{[}26{]}, Sanapareddy, \emph{et al.} {[}25{]}, Wang, \emph{et al.}
{[}15{]}, Weir, \emph{et al.} {[}21{]}, and Zeller, \emph{et al.}
{[}17{]}. The Zackular {[}34{]} study was not included because the 90
individuals analyzed within the study are contained within the larger
Baxter study {[}13{]}. After sequence processing, all the case samples
for the Kostic study had 100 or less sequences remaining and was
excluded, leaving a total of 14 studies that analysis could be completed
on.

@@

\textbf{\emph{Sequence Processing:}} For the majority of studies, raw
sequences were downloaded from the Sequence Read Archive (SRA)
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata were obtained by searching the respective accession number
of the study at the following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
not have sequences and metadata on the SRA, data was obtained from DBGap
(n = 1, {[}12{]}) and directly from the authors (n = 4,
{[}18,21,25,27{]}). Each study was processed using the mothur (v1.39.3)
software program {[}35{]} and quality filtering utilized the default
methods for both 454 and Illumina based sequencing. If it was not
possible to use the defaults, the stated quality cut-offs, from the
study itself, were used instead. Sequences that were made up of an
artificial combination of two or more different sequences and commonly
known as chimeras were identified and removed using VSEARCH {[}36{]}
before \emph{de novo} OTU clustering at 97\% similarity was completed
using the OptiClust algorithm {[}37{]}.

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R (v3.4.3) software package {[}38{]}.
For OTU richness, evenness, and Shannon diversity analysis, values were
power transformed using the rcompanion (v1.11.1) package {[}39{]} and
then Z-score normalized using the car (v2.1.6) package {[}40{]}. Testing
for OTU richness, evenness, and Shannon diversity differences utilized
linear mixed-effect models created using the lme4 (v1.1.15) package
{[}41{]} to correct for study, repeat sampling of individuals (tissue
only), and 16S hyper-variable region used. Odds ratios (OR) were
analyzed using both the epiR (v0.9.93) and metafor (v2.0.0) packages
{[}42,43{]} by assessing how many individuals with and without disease
were above and below the overall median value within each specific
study. OR significance testing utilized the chi-squared test. Diversity
differences measured by the Bray-Curtis index utilized the creation of
distance matrix and testing with PERMANOVA executed with the vegan
(v2.4.5) package {[}44{]}. Random Forest models were built using both
the caret (v6.0.78) and randomForest (v4.6.12) packages {[}45,46{]}.
Power analysis and estimations were made using the pwr (v1.2.1) and
statmod (v1.4.30) packages {[}47,48{]}. All figures were created using
both ggplot2 (v2.2.1) and gridExtra (v2.3) packages {[}49,50{]}.

\textbf{\emph{Study Analysis Overview:}} OTU richness, evenness, and
Shannon diversity were first assessed for differences between controls,
adenoma tumors, and carcinoma turmors using both linear mixed-effect
models and ORs. For each individual study the Bray-Curtis index was used
to assess differences between control-adenoma and control-carcinoma
individuals. Next, all common genera were assessed for differences in
ORs for individuals having an adenoma or carcinoma and corrected for
multiple comparisons using the Benjamini-Hochberg method {[}51{]}. We
then built Random Forest models based on all genera, all OTUs, or
significant OR taxa (if any were present after multiple comparison
correction). For both the full genera and significant OR taxa, models
were trained on one study then tested on the remaining studies using
genera-based relative abundances. The OTU-based models were built using
OTU level data and a 10-fold CV over 100 different iterations, based on
random 80/20 splitting of the data, was used to generate a range of
expected AUCs. This process was repeated for every study in the
meta-analysis. Comparisons of the initial trained model AUCs for the
full genera and significant OR taxa were made to the mean AUC generated
from the 100 different 10-fold CV runs of the respective OTU-based
model. Finally, the power of each study was assessed for an effect size
ranging from 1\% to 30\% and an estimated sample size, for these effect
sizes, was generated based on 80\% power. For comparisons in which only
control versus adenoma individuals were made, the carcinoma individuals
were excluded from each respective study. Similarly, for comparisons in
which control versus carcinoma individuals were made the adenoma
individuals were excluded from each respective study. For all analysis
completed fecal and tissue samples were kept separate. Within the tissue
groups the data were further divided between samples from the same
individual (matched) and those from different individuals (unmatched).

@@

\hypertarget{ref-shah_leveraging_2017}{}
11. Shah MS, DeSantis TZ, Weinmaier T, McMurdie PJ, Cope JL, Altrichter
A, et al. Leveraging sequence-based faecal microbial community survey
data to identify a composite biomarker for colorectal cancer. Gut. 2017;

\hypertarget{ref-ahn_human_2013}{}
12. Ahn J, Sinha R, Pei Z, Dominianni C, Wu J, Shi J, et al. Human gut
microbiome and risk for colorectal cancer. Journal of the National
Cancer Institute. 2013;105:1907--11.

\hypertarget{ref-baxter_microbiota-based_2016}{}
13. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
improves the sensitivity of fecal immunochemical test for detecting
colonic lesions. Genome Medicine. 2016;8:37.

\hypertarget{ref-chen_human_2012}{}
14. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
mucosa-associated microbiota in patients with colorectal cancer. PloS
One. 2012;7:e39743.

\hypertarget{ref-wang_structural_2012}{}
15. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
segregation of gut microbiota between colorectal cancer patients and
healthy volunteers. The ISME journal. 2012;6:320--9.

\hypertarget{ref-burns_virulence_2015}{}
16. Burns MB, Lynch J, Starr TK, Knights D, Blekhman R. Virulence genes
are a signature of the microbiome in the colorectal tumor
microenvironment. Genome Medicine. 2015;7:55.

\hypertarget{ref-zeller_potential_2014}{}
17. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
Potential of fecal microbiota for early-stage detection of colorectal
cancer. Molecular Systems Biology. 2014;10:766.

\hypertarget{ref-flemer_tumour-associated_2017}{}
18. Flemer B, Lynch DB, Brown JMR, Jeffery IB, Ryan FJ, Claesson MJ, et
al. Tumour-associated and non-tumour-associated microbiota in colorectal
cancer. Gut. 2017;66:633--43.

\hypertarget{ref-ecoli_Arthur_2014}{}
19. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
McCafferty J, et al. Microbial genomic analysis reveals the essential
role of inflammation in bacteria-induced colorectal cancer. Nature
Communications {[}Internet{]}. Springer Nature; 2014;5:4724. Available
from: \url{https://doi.org/10.1038/ncomms5724}

\hypertarget{ref-strep_Aymeric_2017}{}
20. Aymeric L, Donnadieu F, Mulet C, Merle L du, Nigro G, Saffarian A,
et al. Colorectal cancer specific conditions promoteStreptococcus
gallolyticusgut colonization. Proceedings of the National Academy of
Sciences {[}Internet{]}. Proceedings of the National Academy of
Sciences; 2017;115:E283--91. Available from:
\url{https://doi.org/10.1073/pnas.1715112115}

\hypertarget{ref-weir_stool_2013}{}
21. Weir TL, Manter DK, Sheflin AM, Barnett BA, Heuberger AL, Ryan EP.
Stool microbiome and metabolome differences between colorectal cancer
patients and healthy adults. PloS One. 2013;8:e70803.

\hypertarget{ref-bfrag_Boleij_2014}{}
22. Boleij A, Hechenbleikner EM, Goodwin AC, Badani R, Stein EM, Lazarev
MG, et al. The bacteroides fragilis toxin gene is prevalent in the colon
mucosa of colorectal cancer patients. Clinical Infectious Diseases
{[}Internet{]}. Oxford University Press (OUP); 2014;60:208--15.
Available from: \url{https://doi.org/10.1093/cid/ciu787}

\hypertarget{ref-geng_diversified_2013}{}
23. Geng J, Fan H, Tang X, Zhai H, Zhang Z. Diversified pattern of the
human colorectal cancer microbiome. Gut Pathogens. 2013;5:2.

\hypertarget{ref-dejea_microbiota_2014}{}
24. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
Rossetti BJ, et al. Microbiota organization is a distinct feature of
proximal colorectal cancers. Proceedings of the National Academy of
Sciences of the United States of America. 2014;111:18321--6.

@@

\hypertarget{ref-Flynn_preprint_2017}{}
28. Flynn KJ, Ruffin MT, Turgeon DK, Schloss PD. Spatial variation of
the native colon microbiota in healthy adults. Cold Spring Harbor
Laboratory; 2017; Available from: \url{https://doi.org/10.1101/189886}

\hypertarget{ref-Salter_contamination_2014}{}
29. Salter SJ, Cox MJ, Turek EM, Calus ST, Cookson WO, Moffatt MF, et
al. Reagent and laboratory contamination can critically impact
sequence-based microbiome analyses. BMC Biology {[}Internet{]}. Springer
Nature; 2014;12. Available from:
\url{https://doi.org/10.1186/s12915-014-0087-z}

\hypertarget{ref-brim_microbiome_2013}{}
30. Brim H, Yooseph S, Zoetendal EG, Lee E, Torralbo M, Laiyemo AO, et
al. Microbiome analysis of stool samples from African Americans with
colon polyps. PloS One. 2013;8:e81352.

\hypertarget{ref-keku_gastrointestinal_2015}{}
31. Keku TO, Dulal S, Deveaux A, Jovov B, Han X. The gastrointestinal
microbiota and colorectal cancer. American Journal of Physiology -
Gastrointestinal and Liver Physiology {[}Internet{]}. 2015 {[}cited 2017
Oct 30{]};308:G351--63. Available from:
\url{http://ajpgi.physiology.org/lookup/doi/10.1152/ajpgi.00360.2012}

>>>>

Colorectal cancer (CRC) is a growing world-wide health problem in which
the microbiota has been hypothesized to have a role in disease
progression {[}1,2{]}. Numerous studies using murine models of CRC have
shown the importance of both individual microbes {[}3--7{]} and the
overall community {[}8--10{]} in tumorigenesis. Numerous case-control
studies have characterized the microbiota of individuals with colonic
adenomas and carcinomas in an attempt to identify biomarkers of disease
progression {[}6,11--17{]}. Because current CRC screening
recommendations are poorly adhered to due to socioeconomic status, test
invasiveness, and frequency of tests, development and validation of
microbiome-associated biomarkers for CRC progression could further
attempts to develop non-invasive diagnostics {[}18{]}.

Recently, there has been an intense focus on identifying
microbiota-based biomarker yielding a seemingly endless number of
candidate taxa. Some studies point towards mouth-associated genera such
as Fusobacterium, Peptostreptococcus, Parvimonas, and Porphyromonas that
are enriched in people with carcinomas {[}6,11--17{]}. Other studies
have identified members of \emph{Akkermansia}, \emph{Bacteroides},
\emph{Enterococcus}, \emph{Escherichia}, \emph{Klebsiella},
\emph{Mogibacterium}, \emph{Streptococcus}, and \emph{Providencia} are
also associated with carciomas {[}13--15{]}. Additionally,
\emph{Roseburia} has been found in some studies to be more abundant in
people with tumors but in other studies it has been found to be either
less abundant or no different than what is found in subjects with normal
colons {[}14,17,19,20{]}. There are strong results from tissue culture
and murine models that Fusobacterium nucleatum, pks-positive strains of
Escherichia coli, Streptococcus gallolyticus, and an
entertoxin-producing strain of Bacteroides fragilis are important in the
pathogenesis of CRC {[}5,14,21--24{]}. These results point to a
causative role for the microbiota in CRC pathogenesis as well as their
potential as diagnostic biomarkers.

Most studies have focused on identifying biomarkers in patients with
carcinomas but there is a greater clinical need to identify biomarkers
associated with adenomas. Studies focusing on broad scale community
metrics have found that measures such as the total number of Operational
Taxonomic Units (OTUs) are decreased in those with adenomas versus
controls {[}25{]}. Other studies have identified \emph{Acidovorax},
\emph{Bilophila}, \emph{Cloacibacterium}, \emph{Desulfovibrio},
\emph{Helicobacter}, \emph{Lactobacillus}, \emph{Lactococcus},
\emph{Mogibacterium}, and \emph{Pseudomonas} to be enriched in those
with adenomas {[}25--27{]}. There are few genera that are enriched in
patients with adenoma or carcinoma tumors.

Confirming some of these previous findings, a recent meta-analysis found
that 16S rRNA gene sequences from members of the \emph{Akkermansia},
\emph{Fusobacterium}, and \emph{Parvimonas} were fecal biomarkers for
the presence of carcinomas {[}28{]}. Contrary to previous studies they
found sequences similar to members of \emph{Lactobacillus} and
\emph{Ruminococcus} to be enriched in patients with adenoma or carcinoma
relative to those with normal colons {[}12,15,16{]}. In addition, they
found 16S rRNA gene sequences from members of \emph{Haemophilus},
\emph{Methanosphaera}, \emph{Prevotella}, \emph{Succinovibrio} were
enriched in patients with adenoma and \emph{Pantoea} were enriched in
patients with carcinomas. Although this meta-analysis was helpful for
distilling a large number of possible biomarkers, the aggregate number
of samples included in the analysis (n = 509) was smaller than several
larger case-control studies that have since been published {[}12,27{]}

Here we provide an updated meta-analysis using 16S rRNA gene sequence
data from both feces (n = 1737) and colon tissue (492 samples from 350
individuals) from 14 studies {[}11--17,19,20,23,25--27,29{]} {[}Table 1
\& 2{]}. We expand both the breadth and scope of the previous
meta-analysis to investigate whether biomarkers describing the bacterial
community or specific members of the community can more accurately
classify patients as having adenoma or carcinoma. Our results suggest
that the bacterial community changes as disease severity worsens and
that that a subset of the microbial community can be used to diagnose
the presence of carcinoma.

@@

\emph{Targeting the identification of tumor microbial biomarkers within
stool seems logical since it offers an easy and cost-effective way to
stratify risk of disease. The current gold standard for diagnosis, a
colonoscopy, can be time-consuming and is not without risk of
complications. Although stool represents an easy and less invasive way
to assess risk, it is not clear how well this sample reflects adenoma-
and carcinoma- associated microbial communities. Some studies have tried
to assess this in health and disease but are limited by their sample
size {[}17,30{]}. Sampling the microbiota directly associated with colon
tissue may provide clearer answers but is not without their own
limitations. After the colonoscopy bowel prep the bacterial community
sampled may reflect the better adhered microbiota versus the resident
community. Additionally, these samples contain more host DNA,
potentially limiting the types of analysis that can be done. It is well
known that low biomass samples can be very difficult to work with and
results can be study dependent due to the randomness of contamination
{[}31{]}.}

@@

One important caveat to this study is that even though genera associated
with certain species such as \emph{Bacteroides fragilis} and
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} were not
identified, it does not necessarily mean that these specific species are
not important in human CRC {[}22,24{]}. Since we are limited in our
aggregation of the data to the genus level, it is not possible to
clearly delineate which species are contributing to overall disease
progression. Our observations are not inconsistent with the previous
literature on either \emph{Bacteroides fragilis} or \emph{Streptococcus
gallolyticus} subsp. \emph{gallolyticus}. As an example, the stool-based
full community models consistently identified the genus
\emph{Bacteroides}, as well as OTUs that classified as
\emph{Bacteroides}, to be important model components across studies.
This suggests that even though \emph{Bacteroides} may not increase the
OR of individuals having an adenoma or carcinoma and may not vary in
relative abundance, like \emph{Fusobacterium}, it is still important in
CRC. Additionally, \emph{Streptococcus gallolyticus} subsp.
\emph{gallolyticus} is a mouth-associated microbe, and the results from
this study suggest that regardless of sample type, mouth-associated
genera are commonly associated with an increased OR for individuals to
have a carcinoma tumor.

@@

\textbf{\emph{Obtaining Data Sets:}} The studies used for this
meta-analysis were identified through the review articles written by
Keku, \emph{et al.} and Vogtmann, \emph{et al.} {[}32,33{]} and
additional studies not mentioned in the reviews were obtained based on
the authors' knowledge of the literature. Studies that used tissue or
feces as their sample source for 454 or Illumina 16S rRNA gene
sequencing analysis and had data sets with sequences available for
analysis were included. Some studies were excluded because they did not
have publicly available sequences or did not have metadata in which the
authors were able to share. After these filtering steps, the following
studies remained: Ahn, \emph{et al.} {[}11{]}, Baxter, \emph{et al.}
{[}12{]}, Brim, \emph{et al.} {[}29{]}, Burns, \emph{et al.} {[}15{]},
Chen, \emph{et al.} {[}13{]}, Dejea, \emph{et al.} {[}20{]}, Flemer,
\emph{et al.} {[}17{]}, Geng, \emph{et al.} {[}19{]}, Hale, \emph{et
al.} {[}27{]}, Kostic, \emph{et al.} {[}34{]}, Lu, \emph{et al.}
{[}26{]}, Sanapareddy, \emph{et al.} {[}25{]}, Wang, \emph{et al.}
{[}14{]}, Weir, \emph{et al.} {[}23{]}, and Zeller, \emph{et al.}
{[}16{]}. The Zackular {[}35{]} study was not included because the 90
individuals analyzed within the study are contained within the larger
Baxter study {[}12{]}. After sequence processing, all the case samples
for the Kostic study had 100 or less sequences remaining and was
excluded, leaving a total of 14 studies that analysis could be completed
on.

@@

\textbf{\emph{Sequence Processing:}} For the majority of studies, raw
sequences were downloaded from the Sequence Read Archive (SRA)
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata were obtained by searching the respective accession number
of the study at the following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
not have sequences and metadata on the SRA, data was obtained from DBGap
(n = 1, {[}11{]}) and directly from the authors (n = 4,
{[}17,23,25,27{]}). Each study was processed using the mothur (v1.39.3)
software program {[}36{]} and quality filtering utilized the default
methods for both 454 and Illumina based sequencing. If it was not
possible to use the defaults, the stated quality cut-offs, from the
study itself, were used instead. Sequences that were made up of an
artificial combination of two or more different sequences and commonly
known as chimeras were identified and removed using VSEARCH {[}37{]}
before \emph{de novo} OTU clustering at 97\% similarity was completed
using the OptiClust algorithm {[}38{]}.

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R (v3.4.3) software package {[}39{]}.
For OTU richness, evenness, and Shannon diversity analysis, values were
power transformed using the rcompanion (v1.11.1) package {[}40{]} and
then Z-score normalized using the car (v2.1.6) package {[}41{]}. Testing
for OTU richness, evenness, and Shannon diversity differences utilized
linear mixed-effect models created using the lme4 (v1.1.15) package
{[}42{]} to correct for study, repeat sampling of individuals (tissue
only), and 16S hyper-variable region used. Odds ratios (OR) were
analyzed using both the epiR (v0.9.93) and metafor (v2.0.0) packages
{[}43,44{]} by assessing how many individuals with and without disease
were above and below the overall median value within each specific
study. OR significance testing utilized the chi-squared test. Diversity
differences measured by the Bray-Curtis index utilized the creation of
distance matrix and testing with PERMANOVA executed with the vegan
(v2.4.5) package {[}45{]}. Random Forest models were built using both
the caret (v6.0.78) and randomForest (v4.6.12) packages {[}46,47{]}.
Power analysis and estimations were made using the pwr (v1.2.1) and
statmod (v1.4.30) packages {[}48,49{]}. All figures were created using
both ggplot2 (v2.2.1) and gridExtra (v2.3) packages {[}50,51{]}.

\textbf{\emph{Study Analysis Overview:}} OTU richness, evenness, and
Shannon diversity were first assessed for differences between controls,
adenoma tumors, and carcinoma turmors using both linear mixed-effect
models and ORs. For each individual study the Bray-Curtis index was used
to assess differences between control-adenoma and control-carcinoma
individuals. Next, all common genera were assessed for differences in
ORs for individuals having an adenoma or carcinoma and corrected for
multiple comparisons using the Benjamini-Hochberg method {[}52{]}. We
then built Random Forest models based on all genera, all OTUs, or
significant OR taxa (if any were present after multiple comparison
correction). For both the full genera and significant OR taxa, models
were trained on one study then tested on the remaining studies using
genera-based relative abundances. The OTU-based models were built using
OTU level data and a 10-fold CV over 100 different iterations, based on
random 80/20 splitting of the data, was used to generate a range of
expected AUCs. This process was repeated for every study in the
meta-analysis. Comparisons of the initial trained model AUCs for the
full genera and significant OR taxa were made to the mean AUC generated
from the 100 different 10-fold CV runs of the respective OTU-based
model. Finally, the power of each study was assessed for an effect size
ranging from 1\% to 30\% and an estimated sample size, for these effect
sizes, was generated based on 80\% power. For comparisons in which only
control versus adenoma individuals were made, the carcinoma individuals
were excluded from each respective study. Similarly, for comparisons in
which control versus carcinoma individuals were made the adenoma
individuals were excluded from each respective study. For all analysis
completed fecal and tissue samples were kept separate. Within the tissue
groups the data were further divided between samples from the same
individual (matched) and those from different individuals (unmatched).

@@

\hypertarget{ref-ahn_human_2013}{}
11. Ahn J, Sinha R, Pei Z, Dominianni C, Wu J, Shi J, et al. Human gut
microbiome and risk for colorectal cancer. Journal of the National
Cancer Institute. 2013;105:1907--11.

\hypertarget{ref-baxter_microbiota-based_2016}{}
12. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
improves the sensitivity of fecal immunochemical test for detecting
colonic lesions. Genome Medicine. 2016;8:37.

\hypertarget{ref-chen_human_2012}{}
13. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
mucosa-associated microbiota in patients with colorectal cancer. PloS
One. 2012;7:e39743.

\hypertarget{ref-wang_structural_2012}{}
14. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
segregation of gut microbiota between colorectal cancer patients and
healthy volunteers. The ISME journal. 2012;6:320--9.

\hypertarget{ref-burns_virulence_2015}{}
15. Burns MB, Lynch J, Starr TK, Knights D, Blekhman R. Virulence genes
are a signature of the microbiome in the colorectal tumor
microenvironment. Genome Medicine. 2015;7:55.

\hypertarget{ref-zeller_potential_2014}{}
16. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
Potential of fecal microbiota for early-stage detection of colorectal
cancer. Molecular Systems Biology. 2014;10:766.

\hypertarget{ref-flemer_tumour-associated_2017}{}
17. Flemer B, Lynch DB, Brown JMR, Jeffery IB, Ryan FJ, Claesson MJ, et
al. Tumour-associated and non-tumour-associated microbiota in colorectal
cancer. Gut. 2017;66:633--43.

\hypertarget{ref-GimenoGarca2012}{}
18. García AZG. Factors influencing colorectal cancer screening
participation. Gastroenterology Research and Practice {[}Internet{]}.
Hindawi Limited; 2012;2012:1--8. Available from:
\url{https://doi.org/10.1155/2012/483417}

\hypertarget{ref-geng_diversified_2013}{}
19. Geng J, Fan H, Tang X, Zhai H, Zhang Z. Diversified pattern of the
human colorectal cancer microbiome. Gut Pathogens. 2013;5:2.

\hypertarget{ref-dejea_microbiota_2014}{}
20. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
Rossetti BJ, et al. Microbiota organization is a distinct feature of
proximal colorectal cancers. Proceedings of the National Academy of
Sciences of the United States of America. 2014;111:18321--6.

\hypertarget{ref-ecoli_Arthur_2014}{}
21. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
McCafferty J, et al. Microbial genomic analysis reveals the essential
role of inflammation in bacteria-induced colorectal cancer. Nature
Communications {[}Internet{]}. Springer Nature; 2014;5:4724. Available
from: \url{https://doi.org/10.1038/ncomms5724}

\hypertarget{ref-strep_Aymeric_2017}{}
22. Aymeric L, Donnadieu F, Mulet C, Merle L du, Nigro G, Saffarian A,
et al. Colorectal cancer specific conditions promoteStreptococcus
gallolyticusgut colonization. Proceedings of the National Academy of
Sciences {[}Internet{]}. Proceedings of the National Academy of
Sciences; 2017;115:E283--91. Available from:
\url{https://doi.org/10.1073/pnas.1715112115}

\hypertarget{ref-weir_stool_2013}{}
23. Weir TL, Manter DK, Sheflin AM, Barnett BA, Heuberger AL, Ryan EP.
Stool microbiome and metabolome differences between colorectal cancer
patients and healthy adults. PloS One. 2013;8:e70803.

\hypertarget{ref-bfrag_Boleij_2014}{}
24. Boleij A, Hechenbleikner EM, Goodwin AC, Badani R, Stein EM, Lazarev
MG, et al. The bacteroides fragilis toxin gene is prevalent in the colon
mucosa of colorectal cancer patients. Clinical Infectious Diseases
{[}Internet{]}. Oxford University Press (OUP); 2014;60:208--15.
Available from: \url{https://doi.org/10.1093/cid/ciu787}

@@

\hypertarget{ref-shah_leveraging_2017}{}
28. Shah MS, DeSantis TZ, Weinmaier T, McMurdie PJ, Cope JL, Altrichter
A, et al. Leveraging sequence-based faecal microbial community survey
data to identify a composite biomarker for colorectal cancer. Gut. 2017;

\hypertarget{ref-brim_microbiome_2013}{}
29. Brim H, Yooseph S, Zoetendal EG, Lee E, Torralbo M, Laiyemo AO, et
al. Microbiome analysis of stool samples from African Americans with
colon polyps. PloS One. 2013;8:e81352.

\hypertarget{ref-Flynn_preprint_2017}{}
30. Flynn KJ, Ruffin MT, Turgeon DK, Schloss PD. Spatial variation of
the native colon microbiota in healthy adults. Cold Spring Harbor
Laboratory; 2017; Available from: \url{https://doi.org/10.1101/189886}

\hypertarget{ref-Salter_contamination_2014}{}
31. Salter SJ, Cox MJ, Turek EM, Calus ST, Cookson WO, Moffatt MF, et
al. Reagent and laboratory contamination can critically impact
sequence-based microbiome analyses. BMC Biology {[}Internet{]}. Springer
Nature; 2014;12. Available from:
\url{https://doi.org/10.1186/s12915-014-0087-z}

\hypertarget{ref-keku_gastrointestinal_2015}{}
32. Keku TO, Dulal S, Deveaux A, Jovov B, Han X. The gastrointestinal
microbiota and colorectal cancer. American Journal of Physiology -
Gastrointestinal and Liver Physiology {[}Internet{]}. 2015 {[}cited 2017
Oct 30{]};308:G351--63. Available from:
\url{http://ajpgi.physiology.org/lookup/doi/10.1152/ajpgi.00360.2012}
